Respiratory Disease Outbreaks in Toronto Pre, During, and Post Covid-19 Pandemic 

### 2025-05-13 paper/paper.qmd
got graphs & tables to work, still need reformatting

@@

*insert table showing first few lines of raw data

@@

*** insert table

Below are the summary statistics for the cleaned data:

*** insert summary statistics

@@

Total & respiratory & Coronavirus:
*** insert graph

@@

Composition of Respiratory Illness:
*** insert graph 
Covid / Secondary Disease / Other & Unknown

@@

Parainfluenza, RSV, Metapneumovirus, and Rhinovirus:
Each has specifics, see general trend of lower number & some w/ slight decrease reported in pre-pandemic & pandemic, with a subsequent increase post.

>>>>

execute:
  echo: false

@@

Below is an example of the raw data from 2019:

import polars as pl
import matplotlib.pyplot as plt

raw_data = pl.read_csv("../data/01-raw_data/ob_report_2019.csv")
raw_data.head(5)

```

@@

```{python}
#| label: disease-count-preview
#| fig-cap: "A line plot on a polar axis"

import polars as pl
import matplotlib.pyplot as plt

raw_data = pl.read_csv("../data/02-analysis_data/disease_count.csv")
raw_data = raw_data.head(5)
raw_data
```
```{python}
#| label: yearly-count-preview
#| fig-cap: "A line plot on a polar axis"

import polars as pl
import matplotlib.pyplot as plt

raw_data = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
raw_data = raw_data.head(5)
raw_data
```

Below are the summary statistics for the cleaned yearly count data:

```{python}
#| label: yearly-count-summary-statistics
#| fig-cap: "A line plot on a polar axis"

import polars as pl
import matplotlib.pyplot as plt

raw_data = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
summary_data = raw_data.describe()
summary_data
```

@@

Total & respiratory & Coronavirus:
```{python}
#| label: all-graph
#| fig-cap: "A line plot on a polar axis"

import polars as pl
import matplotlib.pyplot as plt
import numpy as np

t = np.arange(2019, 2025, 1)
KEYS = ["Coronavirus", "Respiratory", "Total", "Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
raw_data = df.to_dict()
for key in KEYS:
  plt.plot(t, raw_data[key], marker = 'o', label = key)

plt.title("Coronavirus, Respiratory & Total Outbreaks")
plt.xlabel("Year")
plt.ylabel("Outbreak Count")
plt.grid()

plt.legend()
plt.show()
```

@@

Composition of Respiratory Illness:
*** insert graph 
```{python}
#| label: composition-chart
#| fig-cap: "A line plot on a polar axis"

### Workspace setup ####
import numpy as np
import polars as pl
import matplotlib.pyplot as plt

#### Read data ####
df = pl.read_csv("../data/02-analysis_data/disease_count.csv")
disease_comp = df.to_dict()

### Plot data ####
t = np.arange(2019, 2025, 1)

KEYS = ["Coronavirus", "Other", "Unknown"]

for key in KEYS:
  for i in range(6):
    disease_comp[key][i] = 100*disease_comp[key][i]/disease_comp["Total Agents"][i]

corona = np.array(disease_comp["Coronavirus"])
other = np.array(disease_comp["Other"])
unknown = np.array(disease_comp["Unknown"])

ind = np.arange(6)    
width = 0.35

p1 = plt.bar(ind, corona, width, color='#d62728', )
p2 = plt.bar(ind, other, width,  bottom=corona)
p3 = plt.bar(ind, unknown, width,  bottom=corona+other)

plt.ylabel('Percentage of Causative Agents')
plt.title('Composition of Respiratory Disease Outbreaks')
plt.xticks(ind, ('2019', '2020', '2021', '2022', '2023', '2024'))
plt.yticks(np.arange(0, 110, 10))
plt.legend((p1[0], p2[0], p3[0]), ('Coronavirus', 'Other', "Unknown"))

plt.show()
```

Covid / Secondary Disease / Other & Unknown

@@

```{python}
#| label: no-corona-graph
#| fig-cap: "A line plot on a polar axis"

import polars as pl
import matplotlib.pyplot as plt
import numpy as np

t = np.arange(2019, 2025, 1)
KEYS = ["Influenza", "Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
raw_data = df.to_dict()
for key in KEYS:
  plt.plot(t, raw_data[key], marker = 'o', label = key)

plt.title("Other Respiratory Outbreaks")
plt.xlabel("Year")
plt.ylabel("Outbreak Count")
plt.grid()

plt.legend()
plt.show()
```

Parainfluenza, RSV, Metapneumovirus, and Rhinovirus:
```{python}
#| label: no-influenza-graph
#| fig-cap: "A line plot on a polar axis"

import polars as pl
import matplotlib.pyplot as plt
import numpy as np

t = np.arange(2019, 2025, 1)
KEYS = ["Syncytial Virus", "Metapneumovirus", "Rhinovirus", "Parainfluenza"]
df = pl.read_csv("../data/02-analysis_data/yearly_disease_count.csv")
raw_data = df.to_dict()
for key in KEYS:
  plt.plot(t, raw_data[key], marker = 'o', label = key)

plt.title("Other Respiratory Outbreaks")
plt.xlabel("Year")
plt.ylabel("Outbreak Count")
plt.grid()

plt.legend()
plt.show()
```

Each has specifics, see general trend of lower number & some w/ slight decrease reported in pre-pandemic & pandemic, with a subsequent increase post.
